A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Response-to-Reviewers-1.Rmd

I have the same suggestions for Figure 3 as I did for Figure 1.  The resolution is poor, and the mainland should be much larger and the center of the map frames.

> *Done.*

Figure 4: Are the correlations statistically significant? Please provide p-values.

>*The purpose of the daily correlation analysis in Figure 4 is to see how much day to day variability there is in the statistical associations with incidence. The p-values of these correlation coefficients are somewhat misleading because of the smaller samples used in their calculation. The information from this Figure helped us decide which variables to constrain in the SUR systems. At that point, the significance of the correlation is given by the p-values of the coefficients of the model. Instead of reporting the p-values for the daily correlations, we added a sentence to the effect that overall for the period  under examination, the pairwise correlations  between the independent variables and incidence are significant at $p<0.05$, with the exception of the three sunshine variables. More details concerning the correlation tests can be found in the R markdown document in the chunk `correlation-tests`.*

@@

> *Typo fixed.*

Conclusion

Page 24, Lines 43-44: “Thirdly, all environmental data are based on a single station in a province”.  This is a bit concerning since a province is essentially similar to a state.  I suppose the justification that you selected a single station based on population distribution is fair, but an interpolation could have improved the accuracy and precision of the environmental variables.

> *Many thanks for this comment. We were also curious about using a summary of several stations for the environmental variables. To test this, we interpolated the environmental variables using a k-points algorithm. When using these variables, instead of seeing improvements on accuracy and precision, the goodness-of-fit of the models declines, the intervals of confidence become wider, and consequently several coefficients lose significance. It seems to us that the interpolated variables introduce undesirable noise: on the one hand, they smooth the environmental variables over the province, but on the other hand they also include stations in places that are not necessarily informative (such as airports, up in Sierra Nevada, etc.) Having selected the stations manually seems to have kept this information "crisp". We recognize the possibility that a different interpolation algorithm might improve this, and recommend this as a topic for future research. A proper comparison of interpolation algorithms would involve experimenting with several methods to derive several environmental variables for a period of 51 days. For the purpose of this paper, we have recreated the data package to include both datasets (single station and average of stations). In the paper we discuss only the results of the first.*

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Proof-reading the morning after.

Whether these results from laboratory experiments will hold in general remains uncertain. At a global scale, de Ángel Solá et al. [-@deangel2020weathering] see less risk in the Caribean Basin; however, Coelho et al. [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate. Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al. [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States. Yao et al. [-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19. This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al. 2016; Dąbrowska-Leonik et al. 2018; Dinlen et al. 2016; Mathyssen et al. 2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).

In addition to the environmental variables above, we also consider a set of potential socio-economic and demographic confounders.

The first variable that we consider is GDP per capita. Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}. The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated]. In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect. In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic. Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional]. Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown. On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.

@@

The last variable that we consider as a control is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g. metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

@@

The first reported case of COVID-19 in Spain was on January 31st, 2020, when a German tourist in the Canary Islands tested positive for the virus. However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia). In a short period of time, as testing started to ramp up, it became clear that an outbreak was flaring. By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic. This declaration marked a turning point for the public in Spain too. As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe after Italy. In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th. As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g. food, health) and some economic subsectors of industry and construction. A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents. The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed. During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease. Lauer et al. [-@Lauer2020incubation] report for the case of COVID-19 a median incubation period of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The vast majority of cases (95%) develop symptoms between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we judge it best to use lagged values of the environmental variables. We test different time lags as follows. We consider a lagged 8-day average, from date-minus-12 to date-minus-5 days (hereafter *lag8*). Secondly, we consider a lagged 11-day average, from date-minus-12 to date-minus-2 days (hereafter *lag11*). Finally, to account for the likely duration of incubation, we consider a lagged 11-day *weighted* average, from date-minus-12 to date-minus-2 days (hereafter *lag11w*). The weights for this variable are calculated using the parameters of the log-normal distribution reported by Lauer et al. [-@Lauer2020incubation], i.e., a log-mean of 1.621 and a log-standard deviation of 0.418. With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194.

@@

It can be seen in the figure that temperature (in its three forms) has the highest simple correlation with incidence of COVID-19. After temperature, GDP per capita has the highest positive correlation with the dependent variable. The distribution of these correlations is also quite tight over the 30-day period of the study. Hours of sunshine tends to have a moderately high correlation with incidence of COVID-19, however the distribution of these correlations is more spread, and in some cases strays into negative values. A similar thing happens with humidity, which also tends to display more day to day variation in the correlation with the dependent variable. The percentage of older adults shows a relatively tight distribution of day-to-day correlations, and is negative. Population density, in contrast, tends to be negative, but is relatively spread, and on some days, the simple correlation between density and incidence of COVID-19 is weakly positive. Overall for the period under examination, the pairwise correlations  between these variables and incidence are significant at $p<0.05$, with the exception of the three $Sunshine_hours$ variables.

>>>>

Whether these results from laboratory experiments will hold in general remains uncertain. At a global scale, de Ángel Solá et al. [-@deangel2020weathering] see less risk in the Caribean Basin; on the other hand, Coelho et al. [@Coelho2020exponential] warn that at least during the exponential phase, expansion of the virus is not driven by climate. Similarly, whereas Araujo and Naimi [-@Araujo2020spread] argue that spread of SARS-CoV2 will likely be constrained by climate, Harbert et al. [-@Harbert2020spatial] remain unconvinced that spatial modelling can currently discriminate the distribution of the disease on the basis of climate, at least in the United States. Yao et al. [-@Yao2020association], examined data from China and came to the conclusion that neither temperature nor ultraviolet indices had an association with transmission of COVID-19. This is despite previous research that has linked less exposure to UVB radiation to higher prevalence and severity of acute respiratory tract infections (Zittermann et al. 2016; Dąbrowska-Leonik et al. 2018; Dinlen et al. 2016; Mathyssen et al. 2017; Esposito and Lelii 2015; Jat 2017; Moriyama, Hugentobler, and Iwasaki 2020).

In addition to the environmental variables above, it is important to account for potential socio-economic and demographic confounders.

In this respect, the first variable that we consider is GDP per capita. Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}. The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated]. In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect. In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic. Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional]. Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown. On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.

@@

The last variable that we consider as a control is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, but only five provinces have higher order systems of mass mobility (e.g. metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

@@

The first reported case of COVID-19 in Spain was on January 31st, 2020, when a German tourist in the Canary Islands tested positive for the virus. After this case, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia). In a short period of time, as testing started to ramp up, it became clear that an outbreak was flaring. By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic. This declaration marked a turning point for the public in Spain too. As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe after Italy. In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th. As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g. food, health) and some economic subsectors of industry and construction. A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents. The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed. During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease. Lauer et al. [-@Lauer2020incubation] report for the case of COVID-19 a median incubation period of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The vast majority of cases (95%) develop symptoms between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we judge it best to use lagged values of the environmental variables. We test different time lags as follows. We consider a lagged 8-day average, from date-minus-12 to date-minus-5 days (hereafter *lag8*). Secondly, we consider a lagged 11-day average, from date-minus-12 to date-minus-2 days (hereafter *lag11*). Finally, to account for the likely duration of incubation, we consider a lagged 11-day *weighted* average, from date-minus-12 to date-minus-2 days (hereafter *lag11w*). The weights for this variable are calculated using the parameters of the log-normal distribution reported by Lauer et al. [-@Lauer2020incubation], i.e., a log-mean of 1.621 and a log-standard deviation of 0.418. With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194. These weights have the effect of changing the contribution of daily values to the lagged moving average. For instance, the temperature at date-minus-4-days is weighted more heavily than the temperature at date-minus-10-days, as a closer approximation of the conditions at the most likely time of contagion before the disease became manifest.

@@

It can be seen in the figure that temperature (in its three forms) has the highest simple correlation with incidence of COVID-19. After temperature, GDP per capita has the highest positive correlation with the dependent variable. The distribution of these correlations is also quite tight over the 30-day period of the study. Hours of sunshine tends to have a moderately high correlation with incidence of COVID-19, but the distribution of these correlations is more spread, and in some cases strays into negative values. A similar thing happens with humidity, which also tends to display more day to day variation in the correlation with the dependent variable. The percentage of older adults shows a relatively tight distribution of day-to-day correlations, and is negative. Population density, in contrast, tends to be negative, but is relatively spread, and on some days, the simple correlation between density and incidence of COVID-19 is weakly positive. Overall for the period under examination, the pairwise correlations  between these variables and incidence are significant at $p<0.05$, with the exception of the three $Sunshine_hours$ variables.
